---
title: "Lab 10 homework"
author: Stanisław Antonowicz
date: 22/05/2021
output: html_notebook
---

# Homework 1
## Analyze data from *Zheng et al., 2017* similarly to the PBMC3k.

I downloaded `fresh_68k_pbmc_donor_a_filtered_gene_bc_matrices.tar.gz` file and uncompressed it.

### Loading libraries, downloading data

```{r}
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
```
```{r}


# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = "/Users/ant/studia/biological-systems-modelling/lab10/filtered_matrices_mex/hg19/")
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc68k", min.cells = 3, min.features = 200)
pbmc
```
### Computing an additional cell-level metric: a percentage of reads that map to the mitochondrial genome
A theoretical justification is that
> Low-quality / dying cells often exhibit extensive mitochnodrial contamination.
```{r}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```
### Visualizing different metrics, filtering the data
I guess given how many cells are there the plots look worse than in the Seurat tutorial.
```{r}
vlnplot <- VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
vlnplot
```
Maybe `FeatureScatter` will look better.
```{r}
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```
It looks really similar to the one in Seurat tutorial. I will not change their recommended cutoff values.

```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```
### Normalizing the data
With the default parameters.
```{r}
pbmc <- NormalizeData(pbmc)
```
### Selecting highly variable genes
As advised in the Seurat tutorial, the number of most variable genes kept is set to 2000.

```{r}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)
print(top10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```
Interesting -- PPBP is the most highly variable gene in both datasets. Also, PF4, LYZ and IGLL5 are in the top 10 in both of them.

### Scaling the data

This is really basic -- subtract the mean, divide by variance.

This eats all of the RAM (on a 16GB machine) and then crashes when done on all genes, so doing it the default way -- on the most highly expressed ones. The tutorial says:
> Your PCA and clustering results will be unaffected.

I think this is the most important part, so I'm deciding to go with the less resource-intensive version.
```{r}
pbmc <- ScaleData(pbmc)
```
### Running PCA, visualizing
Sadly without a heatmap, as it requires scaling the whole dataset.
```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```
```{r}
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```
```{r}
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```
```{r}
DimPlot(pbmc, reduction = "pca")
```
### Determining dataset's dimensionality

First trying with JackStraw, although it might take really long.
```{r}
pbmc <- JackStraw(pbmc, num.replicate = 100)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20)
```
```{r}
JackStrawPlot(pbmc, dims = 1:20)
```
```{r}
ElbowPlot(pbmc)
```
According to JackStraw procedure, all top 20 PCs are significant. Looking at the ElbowPlot, I think after 15 PCs the plot flattens, so I'm sticking to 15 dimensions in the further analyses.
### Clustering the cells
I tweaked the resolution parameter to achieve 10 clusters.
```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:15)
pbmc <- FindClusters(pbmc, resolution = 0.15)
```

### Visualizing the clustering with UMAP
t-SNE takes extremely long to compute, so sticking with UMAP (I think it's better anyway).
```{r}
pbmc <- RunUMAP(pbmc, dims = 1:15)
```

```{r}
saveRDS(pbmc, file = "68k_pbmc.rds")
```
```{r}
DimPlot(pbmc, reduction = "umap")
```

# Homework 2
A little bit of hacking since I don't know a better way.
First I create a different Seurat object for each cluster.
```{r}
subclusters.objects <- list()

for (cluster.number in 0:(length(unique(pbmc[["seurat_clusters"]][[1]])) - 1)) {
  subclusters.objects[[cluster.number + 1]] <- FindSubCluster(pbmc,
                         cluster.number,
                         graph.name = "RNA_snn",
  )
}
pbmc <- subclusters.objects[[1]]  # little hack to make sure sub.cluster field is properly intialized
```
Then I'm copying subcluster values to sub.cluster metadata in the original Seurat object.
```{r}

for (cluster.number in 0:(length(unique(pbmc[["seurat_clusters"]][[1]])) - 1)) {
  pbmc[["sub.cluster"]][
    pbmc[["seurat_clusters"]] == cluster.number
  ] <- subclusters.objects[[cluster.number + 1]][["sub.cluster"]][[1]][
    pbmc[["seurat_clusters"]] == cluster.number
  ]
}
```

### Visualizing subclusters in one UMAP plot.
```{r}
DimPlot(pbmc, reduction = "umap", label = TRUE, group.by = "sub.cluster", label.size = 3)
```
```{r}
subsetted <- SplitObject(pbmc, split.by = "ident")
```
### Visualizing subclusters on separate plots.
```{r}
first.cluster <- subsetted[[1]][["seurat_clusters"]][[1]][1]
plot <- DimPlot(
        subsetted[[1]],
        reduction = "umap",
        label = TRUE,
        group.by = "sub.cluster",
) + ggtitle(paste("Cluster", as.character(first.cluster), sep = " "))
for (i in 2:(length(unique(pbmc[["seurat_clusters"]][[1]])))) {
  cluster <- subsetted[[i]][["seurat_clusters"]][[1]][1]
  subplot <- DimPlot(
          subsetted[[i]],
          reduction = "umap",
          label = TRUE,
          group.by = "sub.cluster",
  ) + ggtitle(paste("Cluster", cluster, sep = " "))
  plot <- plot + subplot
}
plot
```

